---
title: "Biostat Project"
author: "Confidence Atemnkeng Alemajo"
date: "12/1/2022"
output: html_document
---

```{r setup, include=FALSE, eval=FALSE}
pacman::p_load(haven,tidyverse,table1)
```

```{r}
proj<- read_xpt('/Users/atemalem/Desktop/Courses/Fall 2022/Capstone/LLCP2021.XPT ')
```


```{r}
arthritis<-proj %>% 
   select(`_AGE_G`,`_SEX`,`_IMPRACE`,MARITAL,`_EDUCAG`,`_TOTINDA`,EMPLOY1,HAVARTH5,ARTHEXER,ARTHEDU,LMTJOIN3,JOINPAI2)
names(arthritis)<-c('Age','Sex','Race','Marital','Edu','phys','employ','Arthritis','doc_rec','arthedu','jntlmt','pain')
```



```{r}
arthritis<- arthritis %>% 
   mutate(Marital=recode_factor(Marital,
                               '1'= 'Married',
                               '2'= 'Divorced',
                               '3'= 'Widowed',
                               '4'= 'Separated',
                               '5'= 'Never married',
                               '6'= 'A member of an unmarried couple',
                               '9'= NA_character_)) %>% 
  mutate(phys=recode_factor(phys,
                              '1'='Had physical activity or exercise',
                              '2'='No physical activity or exercise in last 30 days')) %>% 
  mutate(employ=recode_factor(employ,
                              '1'='employed',
                              '2'='employed',
                              '3'='unemployed',
                              '4'='unemployed',
                              '5'='unemployed',
                              '6'='unemployed',
                              '7'='unemployed',
                              '8'='unemployed',
                              '9'=NA_character_)) %>% 
   mutate(Age=recode_factor(Age,
                                '1'= 'Age 18 to 24',
                                '2'= 'Age 25 to 34',
                                '3'= 'Age 35 to 44',  
                                '4'= 'Age 45 to 54',
                                '5'= 'Age 55 to 64', 
                                '6'= 'Age 65 or older')) %>% 
  mutate(Sex=recode_factor(Sex,
                           '1'='male',
                           '2'='female')) %>% 
  mutate(Race=recode_factor(Race,
                            '1'= 'White-Non-Hispanic',
                            '2'= 'Black-Non-Hispanic',
                            '3'= 'Asian, Non-Hispanic',
                            '4'= 'American Indian/Alaskan Native, Non-hispanic',
                            '5'= 'Hispanic',
                            '6'= 'Other race,Non-Hispanic')) %>% 
   mutate(Edu=recode_factor(Edu,
                                 '1'= 'Did not graduate High School',
                                 '2'= 'Graduated High School',
                                 '3'= 'Attended College or Technical School',
                                 '4'= 'Graduated from College or Tech.School',
                                 '9'= NA_character_)) %>% 
  mutate(Arthritis=recode_factor(Arthritis,
                                 '1'= 'Yes',
                                '2'= 'No',
                                '7'= NA_character_,
                                '9'= NA_character_)) %>%
  mutate(doc_rec=recode_factor(doc_rec,
                               '1'= 'Yes',
                                '2'= 'No',
                                '7'= NA_character_,
                                '9'= NA_character_)) %>%
  mutate(arthedu=recode_factor(arthedu,
                               '1'= 'Yes',
                                '2'= 'No',
                                '7'= NA_character_,
                                '9'= NA_character_)) %>%
  mutate(jntlmt=recode_factor(jntlmt,
                              '1'= 'Yes',
                                '2'= 'No',
                                '7'= NA_character_,
                                '9'= NA_character_)) %>% 
  mutate(pain=na_if(pain,77)) %>% 
  mutate(pain=na_if(pain,99)) %>% 
  mutate(pain_severity= cut(pain, breaks=c(0,4,7,11), labels=c('Mild','Moderate','Severe'), include.lowest=TRUE, right=FALSE))

table1(~., data=arthritis)
```





```{r}
arth.clean<-arthritis %>% 
  filter(Arthritis=='Yes') %>% 
  mutate(Age=case_when(Age=='Age 18 to 24'~'18-64 years',
                       Age=='Age 25 to 34'~'18-64 years',
                       Age=='Age 35 to 44'~'18-64 years',
                       Age=='Age 45 to 54'~'18-64 years',
                       Age=='Age 55 to 64'~'18-64 years',
                       Age=='Age 65 or older'~'65 years and older')) %>% 
  mutate(Marital=case_when(Marital=='divorced'~'Single',
                           Marital=='Widowed'~'Single',
                           Marital=='Separated'~'Single',
                           Marital=='Never married'~'Single',
                           Marital=='A member of an unmarried couple'~'Single',
                           Marital=='Married'~'Married')) %>% 
  mutate(Edu=case_when(Edu=='Did not graduate High School'~'Did not/Graduated from high school',
                       Edu=='Graduated High School'~'Did not/Graduated from high school',
                       Edu=='Attended College or Technical School'~'Attended/Graduated from College or Technical school',
                       Edu=='Graduated from College or Tech.School'~'Attended/Graduated from College or Technical school')) %>% 
  mutate(Race=case_when(Race=='White-Non-Hispanic'~'White-Non-Hispanic',
                        Race=='Black-Non-Hispanic'~'Black-Non-Hispanic',
                        Race=='Asian, Non-Hispanic'~'Other race,Non-Hispanic',
                        Race=='American Indian/Alaskan Native, Non-hispanic'~'Other race,Non-Hispanic',
                        Race=='Other race,Non-Hispanic'~'Other race,Non-Hispanic',
                        Race=='Hispanic'~'Hispanic')) %>% 
  drop_na()

```


# Demographic characteristics of Participants
```{r}
library(gtsummary)
gt_table<-tbl_summary(
  arth.clean,
   label = list(Age~"Age(years)",
                Sex~ "Sex",
                Race~ "Race/Ethnicity",
                Marital~"Marital status",
                Edu ~"Level of education",
                employ~"Employment status",
                  jntlmt~"Limitation due to joint symptoms"),
  type =all_dichotomous()~'categorical',
  digits = list(all_categorical() ~ c(0, 1)),  # Number of decimal places for percentages
  missing = 'no',
  include = c(Age,Sex,Race,Marital,Edu,employ,jntlmt) ) %>% 
    bold_labels() %>% 
    modify_caption('**Table1: Demographic Characteristics of Adults Aged 18 years and older with Arthritis in the United States-2021 (N=111212) \ Behavioral risk factor surveillance system data 2021**') 

as_gt(gt_table)

```
 


```{r}

arth.clean<-arth.clean %>% 
        mutate(pain_severity=factor(pain_severity, 
                             levels = c("Mild", "Moderate", "Severe")),
        doc_rec= factor(doc_rec, 
                        levels= c("No", "Yes")),
        arthedu= factor(arthedu, 
                        levels =c("No", "Yes")),
        phys=factor(phys,
                    levels = c('No physical activity or exercise in last 30 days','Had physical activity or exercise')))

```


```{r}
library(nnet)
library(lmtest)
library(car)
```


# pain severity vs doctor suggesting physical activity/exercise(Adjusting for age,sex and race)
```{r}
#visualize the relationship
ggplot(data=arth.clean) + geom_bar(aes(x=pain_severity)) + facet_wrap(~doc_rec) 
 

  plot.docpain<-arth.clean %>% 
    group_by(pain_severity, doc_rec) %>% 
    count() %>% 
    group_by(doc_rec) %>% 
    mutate(percent=100*(n/sum(n))) %>% 
    ggplot(aes(x=doc_rec,fill=pain_severity, y=percent)) +
    geom_col(position = 'dodge',
             color='black')+
    theme_bw() +
    scale_y_continuous(breaks = seq(0,100, by=10)) +
    theme(plot.title = element_text(face = "bold", size = 12),
          axis.text = element_text(colour = "black", size=12),
        axis.title.x = element_text(colour="black",size=12),
        axis.title.y = element_text(colour="black",size=12),
        legend.text = element_text(color="black", size=12),
        legend.title = element_text(color="black", size=12),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x='Doctor/Health professional physical activity recommendations',
         y= 'Percentage in pain severity groups',
         fill= 'Arthritis pain severity',
        title='Participants report of Doctor/Health professional physical activity\
        recommendations by arthritis pain severity',
        subtitle= 'Behavioral Risk Factor Surveillance System Data for 2021(N=111,212)',
                caption = 'BRFSS 2021 data')  
plot.docpain

table1(~doc_rec+arthedu|pain_severity, arth.clean)

# Model
arth_model1<-multinom(pain_severity~doc_rec, data=arth.clean)
broom::tidy(arth_model, exponentiate = TRUE, conf.int = TRUE)
#Adjust for Age,sex and race
arth_model2<-multinom(pain_severity~doc_rec +Age+Sex+Race, data=arth.clean)
broom::tidy(arth_model2, exponentiate = TRUE, conf.int = TRUE)

#compare models
lrtest(arth_model1,arth_model2)


```


# Pain severity vs physical activity in the last 30 days (age,sex,race,edu,marital,arthedu)..Employment a collider so we cannot adjust
```{r}
# visualization of the relationship

  plot.docphys<-arth.clean %>% 
    group_by(pain_severity, phys) %>% 
    count() %>% 
    group_by(phys) %>% 
    mutate(percent=100*(n/sum(n))) %>% 
    ggplot(aes(x=phys,fill=pain_severity, y=percent)) +
    geom_col(position = 'dodge',
             color='black')+
    theme_bw() +
    scale_y_continuous(breaks = seq(0,100, by=10)) +
    theme(plot.title = element_text(face = "bold", size = 12),
          axis.text = element_text(colour = "black", size=12),
        axis.title.x = element_text(colour="black",size=12),
        axis.title.y = element_text(colour="black",size=12),
        legend.text = element_text(color="black", size=12),
        legend.title = element_text(color="black", size=12),
          axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(x='Physical activity in the last 30 days',
         y= 'Percentage in pain severity groups',
         fill= 'Arthritis pain severity',
        title='Participants report of physical activity in the last 30 days\
         by arthritis pain severity',
        subtitle= 'Behavioral Risk Factor Surveillance System Data for 2021(N=111,212)',
                caption = 'BRFSS 2021 data')
plot.docphys


phys_model1<-multinom(pain_severity~phys, data=arth.clean)
broom::tidy(phys_model, exponentiate = TRUE, conf.int = TRUE)
#adjust
phys_model2<-multinom(pain_severity~phys+Age+Sex+Race+Edu+Marital+arthedu, data=arth.clean)
broom::tidy(phys_model2, exponentiate = TRUE, conf.int = TRUE)
# Compare two models 
lrtest(phys_model1,phys_model2)


```

# physical exercise in last 30 days and doctor suggesting physical activity(sex,age,race)
```{r}
# visualize relationship
table1(~doc_rec|phys,data=arth.clean)

phys_suggest1<-glm(phys~doc_rec, data = arth.clean, family=binomial(link = 'logit'))
broom::tidy(phys_suggest, exponentiate = TRUE, conf.int = TRUE)
#Ajust
phys_suggest2<-glm(phys~doc_rec+Sex+Age+Race, data = arth.clean, family=binomial(link = 'logit'))
broom::tidy(phys_suggest2, exponentiate = TRUE, conf.int = TRUE)

#Compare
lrtest(phys_suggest1,phys_suggest2)


```


## Look at assumptions of multicollinearity using the vif function from the car package
```{r}
vif(phys_suggest2)

```


  